# Results
## Relationship Between Drafting High School and College Level Players
We made a barplot that displayed the percentage of draftees by round, with colors indicating high school status at the time of drafting, in order to examine the relationship between draft round and high school status. This barplot analysis shows that a much less number of high school athletes are selected as the first ten rounds go.
```{r barplot, fig.align='center', fig.cap="Percentage of Players by Round and School Level", warning=FALSE, error = TRUE}
#| message: false
library(tidyverse)
library(data.table)

draft_info <- fread('clean_draft.csv')
draft_info <- 
  draft_info %>% 
  mutate(mlb_years = ifelse(mlb_played_first == 0 & mlb_played_last == 0, 0,
                            (mlb_played_last - mlb_played_first) + 1 ) )

# Create a dataframe with state naming data
state_data <- data.frame(names = state.name, abbreviation = tolower(state.abb))

draft_info <- 
  draft_info  
  left_join(state_data, by = c("home_state" = "abbreviation")) |>
  select(-home_state) |>
  rename(home_state = names) |> 
  drop_na(home_state)

players_per_round <- 
  draft_info |> 
  filter(as.integer(pick_round) %in% c(1:10)) |> 
  drop_na(high_school) |> 
  group_by(pick_round) |> 
  summarise(drafted_players = n()) |> 
  ungroup() |> 
  mutate(pick_round = factor(pick_round, levels = c(1:10))) |> 
  arrange(pick_round)

# Subset the data to only players drafted in the first 10 rounds
first_10 <- 
  draft_info |> 
  filter(as.integer(pick_round) %in% c(1:10)) |> 
  drop_na(high_school) |> 
  group_by(pick_round, high_school) |> 
  summarise(num_players = n()) |> 
  arrange(desc(num_players)) |> 
  ungroup() |> 
  inner_join(players_per_round, by = "pick_round") |> 
  mutate(pick_round = factor(pick_round, levels = c(1:10)),
         high_school = as.factor(ifelse(high_school == "Yes", "High School", 
                                        "College") ),
         percent = round(num_players / drafted_players * 100, 2)  )|> 
  arrange(pick_round)

hs_by_round_plot <-
  ggplot(first_10, aes(x = pick_round, y = percent, fill = high_school)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Percent of Draftees", x = "Draft Round", 
       title = "Percent of Draftees by Pick and Round",
       fill = "School Level") +
  theme_minimal() +
  geom_text(aes(label = paste(percent, "%", sep = "")), 
                position = position_stack(vjust = 0.5), size = 4) +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 10))
hs_by_round_plot
```
## How Long to Play First MLB Game
We determined the number of years between a player's first MLB game and the year of their draft in order to examine the number of years prior to a draftee playing their first MLB game. The number of years it takes for a batting draftee to play in Major League Baseball is comparable to that of a pitching draftee, according to an analysis of these two histograms. The majority of players in both datasets play their first game three to five years after being drafted, making them slightly tilted to the right.
```{r histograms, fig.cap = "Histograms of Years before First MLB Game"}
library(gridExtra)

bat_draft_df <- fread('bat_subset.csv')
pitch_draft_df <- fread('pitch_subset.csv')

#Created a new variable for each dataset that gives the number of years it took for a player 
#to play their first MLB game
bat_draft_df$DifferenceYearsDraftFGPlayed <- bat_draft_df$mlb_played_first - bat_draft_df$year
pitch_draft_df$DifferenceYearsDraftFGPlayed <- pitch_draft_df$mlb_played_first - pitch_draft_df$year
plot1 <- ggplot(bat_draft_df, aes(x = DifferenceYearsDraftFGPlayed)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black", alpha = 0.8) +
  labs(title = "Years for Batting Draftee to Play First MLB Game", 
       x = "# of Years", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 10))

plot2 <- ggplot(pitch_draft_df, aes(x = DifferenceYearsDraftFGPlayed)) +
  geom_histogram(binwidth = 0.5, fill = "orange", color = "black", alpha = 0.8) +
  labs(title = "Years for Pitching Draftee to Play First MLB Game", 
       x = "# of Years", y = "Frequency") + 
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 10))
grid.arrange(plot1, plot2, nrow = 1)
```
## Weight vs Height
Additionally, we looked into the correlation between height and weight for both pitching and hitting draftees. In addition to mapping the features of each individual, we also plotted the mean weight/height and carried out a basic linear regression to ascertain the general trend by generating a scatterplot of the two datasets. The mean weight for hitting draftees is marginally lower than the mean weight for pitching draftees, as can be seen when comparing these two plots. Similarly, batting draftees' average height is little lower than that of their pitching counterparts. Compared to the average pitching draftee, the average batting draftee should be a little shorter and lighter.
```{r scatterplots, fig.cap = "Scatterplots of Mean vs. Weight", warning=FALSE}
plot1 <- ggplot(bat_draft_df, aes(x = person_weight, y=person_height)) +
  geom_point(fill="skyblue", shape=21, size=2, alpha = 0.8) +
  geom_hline(yintercept = mean(bat_draft_df$person_height,na.rm=T), 
             color = "red", linetype = "dashed", alpha = 0.8) + 
  geom_vline(xintercept = mean(bat_draft_df$person_weight,na.rm=T), 
             color = "blue", linetype = "dashed", alpha = 0.8) +
  geom_text(aes(x = mean(bat_draft_df$person_weight,na.rm=T) + 1, 
                y = mean(bat_draft_df$person_height,na.rm=T), 
                label = paste("Mean Weight")), 
            color = "blue", hjust = 0, vjust = -15, alpha = 0.8) +  
  geom_text(aes(x = mean(bat_draft_df$person_weight,na.rm=T), 
                y = mean(bat_draft_df$person_height,na.rm=T) - 1, 
                label = paste("Mean Height")), 
            color = "red", hjust = -3, vjust = 0, alpha = 0.8) +  
  geom_smooth(method = "lm", se = FALSE, color = "black", alpha = 0.8) +
  labs(title = "Weight vs. Height (Batters)", 
       x = "Draftee Weight", y = "Draftee Height") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 10))

plot2 <- ggplot(pitch_draft_df, aes(x = person_weight, y=person_height)) +
  geom_point(fill="orange", shape=21, size=2, alpha = 0.8) +
  geom_hline(yintercept = mean(pitch_draft_df$person_height,na.rm=T), 
             color = "red", linetype = "dashed", alpha = 0.8) + 
  geom_vline(xintercept = mean(pitch_draft_df$person_weight,na.rm=T), 
             color = "blue", linetype = "dashed", alpha = 0.8) +
  geom_text(aes(x = mean(pitch_draft_df$person_weight,na.rm=T) + 1, 
                y = mean(pitch_draft_df$person_height,na.rm=T), 
                label = paste("Mean Weight")), 
            color = "blue", hjust = 0, vjust = -16, alpha = 0.8) +  
  geom_text(aes(x = mean(pitch_draft_df$person_weight,na.rm=T), 
                y = mean(pitch_draft_df$person_height,na.rm=T) - 1, 
                label = paste("Mean Height")), 
            color = "red", hjust = -2.8, vjust = 0, alpha = 0.8) +  
  geom_smooth(method = "lm", se = FALSE, color = "black", alpha = 0.8) +
  labs(title = "Weight vs. Height (Pitchers)", 
       x = "Draftee Weight", y = "Draftee Height") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 10))

grid.arrange(plot1, plot2, nrow = 1)
```
## Correlation Heatmap
For both player categories, the following association exists between pick round, year, height, weight, and WAR: As anticipated, a person's height and weight are strongly correlated. There is a slight positive association between a player's height and WAR while they are pitching, but a negative correlation when they are batting. We find that the weight of a batting draftee and WAR do not correlate, but for a pitching draftee, there is a little positive correlation.

```{r correlation,fig.align='center', fig.cap = "Correlation Plots for Pitching and Batting", fig.height=3.5, warning=FALSE}
library(corrplot)

par(mfrow=c(1,2))
# >| full_join(pitch_draft_df)
bat_draft_df1 <- bat_draft_df %>%
  mutate(pick_round=as.numeric(pick_round)) %>% 
  drop_na()
correlation_matrix <- cor(bat_draft_df1[, c(4,5,7,9,10)]) 
corrplot(correlation_matrix, method = "color")
title("Correlation Plot (Batting)",line = 3,adj=0)
#title(sub = "Plot 5: Correlation Plot of Batters", line=4.5)

pitch_draft_df1 <- pitch_draft_df %>%
  mutate(pick_round=as.numeric(pick_round)) %>% 
  drop_na()
correlation_matrix <- cor(pitch_draft_df1[, c(4,5,7,9,10)]) 
corrplot(correlation_matrix, method = "color")
title("Correlation Plot (Pitching)",line = 3,adj=0)
#title(sub = "Plot 6: Correlation Plot of Pitchers", line=4.5)
```